\assignment

### 2017-04-04 project/paper.rnw
About done with the project.

\newcommand\assignment{Stat 534 Project: \\
Extrapolation from Poisson Process Intensity Surface Models}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{April 2, 2017}

@@

par(mar = c(0, 0, 1, 3))
plot(intense.im, main = 'True Intensity Surface',
     border = NA, zlim = c(100, 300), las = 2)
plot(setminus.owin(owin(c(1556870, 1564495), c(535421, 541130),
                        unitname = c('foot', 'feet')),
     site_window), col = 'white', border = 'white', lwd = 2, add = TRUE)
mtext('Anomalies per Acre', 4, line = 2)
@

@@

<<polypred1>>=
# Get predictions and SEs.
ex_p2 <- predict(ex_fit2, window = site_window, se = TRUE)
ex_p3 <- predict(ex_fit3, window = site_window, se = TRUE)
ex_p4 <- predict(ex_fit4, window = site_window, se = TRUE)
ex_p5 <- predict(ex_fit5, window = site_window, se = TRUE)
ex_p6 <- predict(ex_fit6, window = site_window, se = TRUE)
ex_p7 <- predict(ex_fit7, window = site_window, se = TRUE)
ex_p8 <- predict(ex_fit8, window = site_window, se = TRUE)
ex_p9 <- predict(ex_fit9, window = site_window, se = TRUE)
ex_p10 <- predict(ex_fit10, window = site_window, se = TRUE)
ex_p11 <- predict(ex_fit11, window = site_window, se = TRUE)
ex_p12 <- predict(ex_fit12, window = site_window, se = TRUE)

@@

The kernel density estimate is computed using the \texttt{spatstat}'s default
``simple rule of thumb'' to select the smoothing bandwidth. Because the
anomalies located more than three feet from a transect were not observed, I
would expect the kernel density estimate to show ridges of high intensity
along the transects. As it turns out, the default bandwith is large enough to
smooth across multiple transects so the ridges do not appear. However, the
kernel density estimate is biased low because not all anomalies are observed;
I divide the estimate by the proportion of the area observed as an approximate
bias correction and the result is a surprisingly reasonable intensity map.

<<polyplot1, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = 'Predicted intensity surfaces for eleven polynomial models, and a kernel density estimate for comparision.'>>=
par(mfrow = c(4, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)

# Multiply predictions by 43560 to convert from anomalies per square foot
# to anomalies per acre.
plot(43560 * ex_p2$estimate, zlim = c(0, 300), main = '2nd Degree')
plot(43560 * ex_p3$estimate, zlim = c(0, 300), main = '3rd Degree')
plot(43560 * ex_p4$estimate, zlim = c(0, 300), main = '4th Degree')
plot(43560 * ex_p5$estimate, zlim = c(0, 300), main = '5th Degree')
plot(43560 * ex_p6$estimate, zlim = c(0, 300), main = '6th Degree')
plot(43560 * ex_p7$estimate, zlim = c(0, 300), main = '7th Degree')
plot(43560 * ex_p8$estimate, zlim = c(0, 300), main = '8th Degree')
plot(43560 * ex_p9$estimate, zlim = c(0, 300), main = '9th Degree')
plot(43560 * ex_p10$estimate, zlim = c(0, 300), main = '10th Degree')
plot(43560 * ex_p11$estimate, zlim = c(0, 300), main = '11th Degree')
plot(43560 * ex_p12$estimate, zlim = c(0, 300), main = '12th Degree')
plot(43560 * ex_kde, zlim = c(0, 500), main = 'Kernel Density')
@

@@

<<polyse1, fig.width = 6, fig.height = 4, fig.pos = 't', fig.cap = 'Log standard errors for six of the polynomial models.'>>=
par(mfrow = c(2, 3), mar = c(1, 0, 2, 3), las = 2, cex = 1)
plot(log(43560 * ex_p2$se), main = '2nd Degree\nlog(SE)')
plot(log(43560 * ex_p4$se), main = '4th Degree\nlog(SE)')
plot(log(43560 * ex_p6$se), main = '6th Degree\nlog(SE)')
plot(log(43560 * ex_p8$se), main = '8th Degree\nlog(SE)')
plot(log(43560 * ex_p10$se), main = '10th Degree\nlog(SE)')
plot(log(43560 * ex_p12$se), main = '12th Degree\nlog(SE)')
@

@@

\texttt{plot} doesn't work, and \texttt{predict} doesn't work on the original
window

<<dummy>>=
par(mfrow = c(1, 2), mar = c(1, 0, 1, 0), cex = 1)
plot(ex_fit1$Q$dummy, pch = '.', border = NA,
     main = 'Default Dummy Points')
plot(site_window, add = TRUE)
plot(ex_Q$dummy, pch = '.', border = NA,
     main = 'Manual Dummy Points')
plot(site_window, add = TRUE)
@

\section{Discussion and Conclusion}

smallish SEs outside observed region because of faith in the model form --
need model checking

What geostatisticians and ordnance engineers will notice is that that SEs
don't reflect added uncertainty about unobserved regions.
Note kriging accounts for lack of model fit by autocorrelated errors --
prediction SEs increase far from obs because of lack of info about local
deviations from mean. For Poisson process, the events are independent given
the intensity so local variation is not helpful for prediction. Thus we are
heavily reliant on the correctness of the model form and the
representativeness of the observed subregion, so model checking is crucial.
If there is prior uncertainty about the form of the trend in a certain part of
the site, that region must be observed!

how much of the region do you need to observe to trust your model?

>>>>

\newcommand\assignment{Stat 534 Project: \\
Extrapolation from Poisson Process Intensity Surface Models}
\newcommand\myname{Kenny Flagg}
\newcommand\duedate{April 3, 2017}

@@

par(mar = c(0, 0, 1, 3))
plot(intense.im, main = 'True Intensity Surface',
     border = NA, zlim = c(0, 300), las = 2)
plot(setminus.owin(owin(c(1556870, 1564495), c(535421, 541130),
                        unitname = c('foot', 'feet')),
     site_window), col = 'white', border = 'white', lwd = 2, add = TRUE)
mtext('Anomalies per Acre', 4, line = 2)
@

@@

<<polypred1, cache = TRUE, dependson = 'quad1'>>=
# Get predictions and SEs.
ex_p2 <- predict(ex_fit2, window = site_window, se = TRUE)
ex_p3 <- predict(ex_fit3, window = site_window, se = TRUE)
ex_p4 <- predict(ex_fit4, window = site_window, se = TRUE)
ex_p5 <- predict(ex_fit5, window = site_window, se = TRUE)
ex_p6 <- predict(ex_fit6, window = site_window, se = TRUE)
ex_p7 <- predict(ex_fit7, window = site_window, se = TRUE)
ex_p8 <- predict(ex_fit8, window = site_window, se = TRUE)
ex_p9 <- predict(ex_fit9, window = site_window, se = TRUE)
ex_p10 <- predict(ex_fit10, window = site_window, se = TRUE)
ex_p11 <- predict(ex_fit11, window = site_window, se = TRUE)
ex_p12 <- predict(ex_fit12, window = site_window, se = TRUE)

@@

The kernel density estimate is computed using \texttt{spatstat}'s default
``simple rule of thumb'' to select the smoothing bandwidth. Because the
anomalies located more than three feet from a transect were not observed, I
would expect the kernel density estimate to show ridges of high intensity
along the transects. As it turns out, the default bandwith is large enough to
smooth across multiple transects so the ridges do not appear. However, the
kernel density estimate is biased low because not all anomalies are observed;
I divide the estimate by the proportion of the area observed as an approximate
bias correction and the result is a surprisingly reasonable intensity map.

<<polyplot1, fig.width = 6, fig.height = 8, fig.pos = 'p', fig.cap = 'Predicted intensity surfaces for eleven polynomial models, and a kernel density estimate for comparision. Note that the 4th- through 12th-degree polynomial models have the same color scale, while the 2nd- and 3rd-degree models and the kernel density estimate use different color scales for visibility.'>>=
par(mfrow = c(4, 3), mar = c(1, 0, 1.2, 3), las = 2, cex = 1)

# Multiply predictions by 43560 to convert from anomalies per square foot
# to anomalies per acre.
plot(43560 * ex_p2$estimate, main = '2nd Degree')
plot(43560 * ex_p3$estimate, main = '3rd Degree')
plot(43560 * ex_p4$estimate, zlim = c(0, 300), main = '4th Degree')
plot(43560 * ex_p5$estimate, zlim = c(0, 300), main = '5th Degree')
plot(43560 * ex_p6$estimate, zlim = c(0, 300), main = '6th Degree')
plot(43560 * ex_p7$estimate, zlim = c(0, 300), main = '7th Degree')
plot(43560 * ex_p8$estimate, zlim = c(0, 300), main = '8th Degree')
plot(43560 * ex_p9$estimate, zlim = c(0, 300), main = '9th Degree')
plot(43560 * ex_p10$estimate, zlim = c(0, 300), main = '10th Degree')
plot(43560 * ex_p11$estimate, zlim = c(0, 300), main = '11th Degree')
plot(43560 * ex_p12$estimate, zlim = c(0, 300), main = '12th Degree')
plot(43560 * ex_kde, main = 'Kernel Density')
@

@@

<<polyse1, fig.width = 6, fig.height = 4, fig.pos = 'p', fig.cap = 'Log standard errors for six of the polynomial models.'>>=
par(mfrow = c(2, 3), mar = c(1, 0, 2, 3), las = 2, cex = 1)
plot(log(43560 * ex_p2$se), main = '2nd Degree\nlog(SE)')
plot(log(43560 * ex_p4$se), main = '4th Degree\nlog(SE)')
plot(log(43560 * ex_p6$se), main = '6th Degree\nlog(SE)')
plot(log(43560 * ex_p8$se), main = '8th Degree\nlog(SE)')
plot(log(43560 * ex_p10$se), main = '10th Degree\nlog(SE)')
plot(log(43560 * ex_p12$se), main = '12th Degree\nlog(SE)')
@

<<dummy, fig.pos = 'p', fig.cap = 'The default dummy points selected by \\texttt{spatstat} (left) and the 500,000 random uniform dummy points that I use to fit the models.'>>=
par(mfrow = c(1, 2), mar = c(1, 0, 1, 0), cex = 1)
plot(ex_fit1$Q$dummy, pch = '.', border = NA,
     main = 'Default Dummy Points')
plot(site_window, add = TRUE)
plot(ex_Q$dummy, pch = '.', border = NA,
     main = 'Manual Dummy Points')
plot(site_window, add = TRUE)
@

@@

I encountered many difficulties in fitting these models that suggest the
methods and software are not yet developed well enough to make them useful for
working with such sparsely sampled data. Most of these difficulties resulted
from the implementation with \texttt{spatstat}, but several were computational
issues.

The troubles with \texttt{spatstat} largely come from how the package sets up
grids of locations for dummy points and predictions. It finds a bounding box
for the region and places grid cells on a lattice across the bounding box, but
its grids are so coarse that most of my transects do not have cells centered
on them. It attempts to create an \(80 \times 80\) grid of dummy points, but
the only useable points are 160 points on two transects; all others are
discarded~(Figure~\ref{fig:dummy}, left). Attempting to fit models in
\texttt{ppm} using these dummy points results in an error.

The \texttt{ppm} function fits models successfully when using the 500,000
dummy points that I draw from a uniform distribution~(Figure~\ref{fig:dummy},
right). However, the \texttt{plot} method gives an error about infinite
values, while the \texttt{predict} method by default uses the same
\(80 \times 80\) grid used for the default dummy points and makes predictions
only in the 160 grid cells that happen to lie inside the observed region.
Fortunately, the \texttt{predict} function behaves when the whole site is
specified as the prediction window.

The biggest disappoinment of using \texttt{spatstat} to fit these models is
that models with splines cannot be plotted. The \texttt{ppm} functions
estimates the parameters without any apparent trouble, but the \texttt{plot}
and \texttt{predict} functions give errors because they do not know how to
interpret the smooth terms in the model formula.

All of the above difficulties are software issues that could be fixed. I also
found two computational limitations that will not be remedied as easily.
First, even after rescaling the coordinates, \(\mathbf{X}^{T}\mathbf{X}\)
cannot be inverted because of numerical instabilities when \(p \geq 18\). Such
models are more complicated than needed for my example site, but it is
important to note that a limit on polynomial model complexity exists.

Second, choosing dummy points is hard. I initially used 1,000 points, but that
was not enough and the intensity values predicted by those models were in the
tens of thousands. It is not apparent that the 500,000 randomly-selected dummy
points I ultimately used are optimal, either. \textcite{wartonshepherd}
discuss dummy point selection when the entire region of interest is observed,
suggesting that the number of dummy points be increased until the maximized
log-likelihood reaches approximate convergence. They use 86,227 points for
their example. I used trial-and-error to increase the number of dummy points,
fitting a couple of my models using 600,000 and then 1,000,000 dummy points;
the maximized log-likelihood changed by thousands. I settled on 500,000
because that was the smallest number I tried that yielded predicted
intensities of the correct magnitude. With that many points, each model takes
about a minute to fit; models with enough dummy points to accurately
approximate the likelihood would likely take much longer to estimate.

\pagebreak
\section{Discussion and Conclusion}

Spatial point process intensity surface models are an intriguing option for
modeling partially-observed inhomogeneous point processes, and polynomial
models can approximate complex surfaces. However, these models are not
yet ready for application to UXO data because the software is not quite there
(yet) and they require heavy computing power. Several avenues for future work
are available, such as seeking more efficient ways to approximate the
likelihood and compute parameter estimates, and further exploring how to
choose dummy points. Then, the next step might be to study whether familiar
model selection tools like AIC or tests of predictive ability are suitable
assuming the likelihood approximation is successful. Presently, it seems
necessary either to observe more of the site area than covered by typical UXO
sampling plans, or work with a model fit that is a rough approximation at
best. Thus the biggest open question might be how much of the region do you
need to observe to trust your model?
